Ultrafast Electron Microscopes: Design Criteria, Electron Sources, and Column Modeling

### 2013-04-12 inc/conclusions.tex
Some last-minute tweaks

Ultrafast Electron Microscopy is an exciting new field of instrumentation research which promises to open our eyes to the dynamics occurring in nanoscale processes.

My contributions to this field have worked to advance this technology, I have developed a model and implemented it, and I have  both the theoretical (simulation/modeling) and experimental (prototyping and photocathode engineering) fronts.

@@

This thesis expounds on several factors that are key concepts in designing a UEM.
I have used the above model to show that for a given electron pulse, whether oblate (disk-like) or prolate (cigar-like), its initial expansion will be primarily in the same direction as its shorter dimension, i.e., an electron pulse ``wants'' to be spherical.
The model predicts that the use of shorter focal-length magnetic lenses will result in higher fidelity focusing; such lenses limit the impulse delivered to the pulse by reducing the time of flight that the pulse experiences the higher charge densities inherent during focusing.
Additionally, we have seen that oblate pulses are preferred to prolate pulses in terms of focus fidelity.
The model also demonstrates that when properly designed and driven, RF cavities can effectively recompress electron pulses which have expanded longitudinally.
Interestingly, the model suggests that the additional bandwidth generated by space-charge repulsion (especially in higher charge-density pulses) can allow for additional pulse compression even at one-to-one image to object distances.

@@

We have also designed and built a custom laser source capable of generating these electron pulses.
The primary laser radiation source for the studies is a home-built, diode-pumped and thermal-lens-shaped, femtosecond Yb:KGW oscillator~\cite{berger_high-power_2008} delivering 250fs duration pulses at 1047nm and a 63MHz repetition rate.
This 2W average power laser is frequency doubled in a 3mm lithium triborate (LBO) crystal with 50\% efficiency to yield $\sim$200fs green pulses at 523nm (photon energy, $\hbar \omega$ = 2.37eV).
Further doubling with a cylindrical focusing geometry in a 7mm $\beta$-barium borate (BBO) crystal yields 261nm ($\hbar \omega$ = 4.75eV) UV pulses with a duration of $\sim$4ps (due to the 600fs/mm group velocity mismatch between the green and generated UV).
As an ancillary benefit of using an in-house built laser source, the cavity length may be changed; for soliton mode-locked lasers, this length is directly related to the repitition rate.
By tuning the length, we are able to frequency match a high harmonic of the repitition rate and the natural oscillation of the RF cavity.
This is done to synchronize the laser and the RF cavity in an essentially passive manner, obviating the need for expensive electronic devices which would be add deleterious electronic phase noise (jitter) to the system.

@@

Our most promising results have come from excited-state thermionic emission (ESTE) sources, GaSb and InSb.
In this process the laser promotes carriers from the valence band to an upper conduction band close to the vacuum level, reducing the effective work function.
The electron temperature, elevated by the laser-driven photoemission process to $\sim10^3$K, causes the tail of the Fermi distribution to exceed the effective work function and thus emission occurs.
Importantly, the results we observed were inconsistent with predictions until we considered the effective mass of these upper conduction bands.
Not only did this allow our prediction to match the experimental results, it opened our eyes to the possibilities of low effective mass materials, even for photoemission processes other than ESTE.
As I have mentioned earlier, this work is ongoing and is a very exciting avenue for future research.

>>>>

Ultrafast Electron Microscopy is an exciting new field of instrumentation research which promises to open our eyes to the dynamics occurring in nanoscale processes.
Through my contributions to this field, I have worked to advance this new technology.
These include a developing and implementing a model for simulating the dynamics of ultrafast electron pulses in electron microscope systems, deriving necessary design critera for constructing such a system, and laying the theoretical and experimental groundwork towards selecting a useful photocathode for electron pulse generation.
These advances have guided the design and construction of the prototype ultrafast electron microscope system being built at UIC.

@@

This thesis expounds on several factors that are key concepts in designing a UEM.
I have used the above model to show that for a given electron pulse, whether oblate (disk-like) or prolate (cigar-like), its initial expansion will be primarily in the same direction as its shorter dimension, i.e., an electron pulse ``wants'' to be spherical.
The model predicts that the use of shorter focal-length magnetic lenses will result in higher fidelity focusing; such lenses limit the impulse delivered to the pulse by reducing the time of flight that the pulse experiences the higher charge densities inherent during focusing.
Additionally, we have seen that oblate pulses are preferred to prolate pulses in terms of focus fidelity.
The model also demonstrates that when properly designed and driven, RF cavities can effectively recompress electron pulses which have expanded longitudinally.
Interestingly, the model suggests that the additional bandwidth generated by space-charge repulsion (especially in higher charge-density pulses) can allow for additional pulse compression, to durations shorter than expected, even at one-to-one image to object distances.

@@

We have also designed and built a custom laser source capable of generating these electron pulses.
The primary laser radiation source for the studies is a home-built, diode-pumped and thermal-lens-shaped, femtosecond Yb:KGW oscillator~\cite{berger_high-power_2008} delivering 250fs duration pulses at 1047nm and a 63MHz repetition rate.
This 2W average power laser is frequency doubled in a 3mm lithium triborate (LBO) crystal with 50\% efficiency to yield $\sim$200fs green pulses at 523nm (photon energy, $\hbar \omega$ = 2.37eV).
Further doubling with a cylindrical focusing geometry in a 7mm $\beta$-barium borate (BBO) crystal yields 261nm ($\hbar \omega$ = 4.75eV) UV pulses with a duration of $\sim$4ps (due to the 600fs/mm group velocity mismatch between the green and generated UV).
As an ancillary benefit of using an in-house built laser source, the cavity length may be changed; for soliton mode-locked lasers, this length is directly related to the repetition rate.
By tuning the length, one may frequency match a high harmonic of the repetition rate and the natural oscillation of the RF cavity.
As a result, the laser and the RF cavity may be synchronized in an essentially passive manner, obviating the need for expensive electronic devices which would be add deleterious electronic phase noise (jitter) to the system.

@@

The most promising results have come from excited-state thermionic emission (ESTE) sources, GaSb and InSb.
In this process the laser promotes carriers from the valence band to an upper conduction band close to the vacuum level, reducing the effective work function.
The electron temperature, elevated by the laser-driven photoemission process to $\sim10^3$K, causes the tail of the Fermi distribution to exceed the effective work function and thus emission occurs.
Importantly, the results we observed were inconsistent with predictions until we considered the effective mass of these upper conduction bands.
Not only did this allow our prediction to match the experimental results, it opened our eyes to the possibilities of low effective mass materials, even for photoemission processes other than ESTE.
As I have mentioned earlier, this work is ongoing and is a very exciting avenue for future research.

### 2013-04-12 inc/introduction.tex
Some last-minute tweaks

The development of laboratory-scale instrumentation for the direct visualization of fundamental ultrafast (i.e., sub-nanosecond) events with nano-scale (even atomic) spatial resolution is a current active area of research~\cite{king_ultrafast_2005}.
In many areas of physics, chemistry and materials science a space-time resolution in the sub-1nm$\cdot$ps range is ideally required to allow for the observation of the fastest structural events.
A promising approach is to employ ultrashort pulses of electrons generated by a laser-driven 10-100keV photoelectron gun to probe the dynamics of the specimen that has been perturbed by a synchronized laser pulse~\cite{king_ultrafast_2005}.
Ultrafast electron diffraction (UED) or femtosecond electron diffraction (FED)~\cite{srinivasan_ultrafast_2003,williamson_clocking_1997,cao_femtosecond_2003} is an example of this type of technique and has been used successfully to elucidate atomic-scale dynamics in melting~\cite{cao_femtosecond_2003,sciaini_electronic_2009,siwick_atomic-level_2003}, chemical bond breaking~\cite{zewail_laser_1988}, etc.
Direct time-resolved imaging of similar events can be accomplished using transmission electron microscopy (TEM).
To date, such dynamic transmission electron microscopes (DTEMs)~\cite{bostanjoglo_tracing_1980,domer_high-speed_2003} have been developed for both single-shot imaging with $\sim$$10^{5}$nm$\cdot$ps space-time resolution (e.g., 30ns electron pulses and 5-10nm spatial resolution~\cite{lagrange_single-shot_2006,armstrong_practical_2007}) and $\sim$1nm$\cdot$ps space-time resolution ($\sim$100fs with a few nanometer resolution~\cite{park_direct_2009}) in the multi-shot data acquisition mode of operation.
Both of these instruments employ retro-fitted standard electron microscope columns and are limited in performance mainly by space-charge effects.

@@

In Chapter \ref{chap:ms_model}, I will introduce the of the work of Michalik and Sipe~\cite{michalik_analytic_2006,michalik_erratum:_2008}, describing free-space electron bunch propagation in an mean-field self-similar Analytic Gaussian (AG) pulse treatment.
For free-space propagation, the AG model of charge bunch dynamics has already been shown to be very consistent with full Monte Carlo (i.e., particle tracking) simulations for a wide variety of electron pulse shapes~\cite{michalik_analytic_2006,michalik_evolution_2009}, including the uniform ellipsoid~\cite{luiten_how_2004}.
Due primarily to the versatility of the AG model which results from its use of transverse and longitudinal pulse position and momentum variances, the AG approach is applicable to both the single electron per pulse limit~\cite{lobastov_four-dimensional_2005}, where momentum variances determine the pulse evolution and the model is exact (obeying Gaussian optics), and the high charge density limit in which space-charge effects dominate~\cite{luiten_how_2004,siwick_ultrafast_2002,cao_femtosecond_2003}.
Unfortunately, the model lacks the ability to model the dynamics of electron pulses under influence of external forces; this prevents the AG model from being used to simulate the behavior of electron pulses in realistic microscope columns.
Additionally, to be useful for UEM, for which initial pulse conditions are of notable importance for the performance of the instrument, a realistic set of intial conditions is essential for accurately modeling the pulse dynamics.

In Chapter \ref{chap:extension}, I present some optimizations to the AG model, and develop useful initial conditions representing single-photon photoemission from a typical metal photocathode.
I then present my extensions to the Michalik and Sipe AG model which permit the inclusion of generic linear external forces, such as those due to perfect electron-optical elements in a microscope column.
In order to adapt this extended AG model for simulating pulses in a realistic UEM column, I then present specific contributions that arise from magnetic lenses, RF cavities and DC accelerators.
The resulting computationally efficient propagation analysis can then be used to model, and hence design, UEM columns and UED systems in a straightforward manner.
Note that the analyses presented are performed in the non-relativistic limit, which is a reasonable approximation for the typical 20-200keV electron energies employed in UED and electron microscopy.

In Chapter \ref{chap:model_results}, I apply the extended AG model to draw some initial insight into the dynamics of ultrafast electron pulses traversing a UEM column.
One of the conclusions that I will make is that oblate, or ``disk-like'', electron pulses will be generally preferred to prolate ``cigar-like;'' for UEM, the electron pulses are likely to be oblate for reasonably sized incident laser spot sizes.
Then in Chapter \ref{chap:considerations}, I apply some well known results from the field of electron microcopy, namely the Rose criterion~\cite{rose_television_1948}, an ultrafast analog to the Child-Langmuir law~\cite{child_discharge_1911,langmuir_effect_1923,valfells_effects_2002}, and the conservation of transverse emittance~\cite{jensen_emittance_2010} (each of which will be introduced) to arrive at two design goals for building a UEM which will produce high-resolution images.
First, that the electron pulses should be emitted from a large-area source, and second, that every effort must be made to reduce the initial rms transverse momentum (i.e., the emission cone angle).

>>>>

The development of laboratory-scale instrumentation for the direct visualization of fundamental ultrafast (i.e., sub-nanosecond) events with nano-scale (even atomic) spatial resolution is a current active area of research~\cite{king_ultrafast_2005}.
In many areas of physics, chemistry and materials science a space-time resolution in the sub-1nm$\cdot$ps range is ideally required to allow for the observation of the fastest structural events.
A promising approach is to employ ultrashort pulses of electrons generated by a laser-driven 10-100keV photoelectron gun to probe the dynamics of the specimen that has been perturbed by a synchronized laser pulse~\cite{king_ultrafast_2005}.
Ultrafast electron diffraction (UED) or femtosecond electron diffraction (FED)~\cite{srinivasan_ultrafast_2003,williamson_clocking_1997,cao_femtosecond_2003} is an example of this type of technique and has been used successfully to elucidate atomic-scale dynamics in melting~\cite{cao_femtosecond_2003,sciaini_electronic_2009,siwick_atomic-level_2003}, chemical bond breaking~\cite{zewail_laser_1988}, etc.
Direct time-resolved imaging of similar events can be accomplished using transmission electron microscopy (TEM).
To date, such dynamic transmission electron microscopes (DTEMs)~\cite{bostanjoglo_tracing_1980,domer_high-speed_2003} have been developed for both single-shot imaging with $\sim$$10^{5}$nm$\cdot$ps space-time resolution (e.g., 30ns electron pulses and 5-10nm spatial resolution~\cite{lagrange_single-shot_2006,armstrong_practical_2007}) and $\sim$1nm$\cdot$ps space-time resolution ($\sim$100fs with a few nanometer resolution~\cite{park_direct_2009}) in the multi-shot data acquisition mode of operation.
Both of these instruments, which operate with vastly different pulse charge densities, employ retro-fitted standard electron microscope columns and are limited in performance mainly by space-charge effects.

@@

In Chapter \ref{chap:ms_model}, I will introduce the of the work of Michalik and Sipe~\cite{michalik_analytic_2006,michalik_erratum:_2008}, describing free-space electron bunch propagation in an mean-field self-similar Analytic Gaussian (AG) pulse treatment.
For free-space propagation, the AG model of charge bunch dynamics has already been shown to be very consistent with full Monte Carlo (i.e., particle tracking) simulations for a wide variety of electron pulse shapes~\cite{michalik_analytic_2006,michalik_evolution_2009}, including the uniform ellipsoid~\cite{luiten_how_2004}.
Due primarily to the versatility of the AG model, which results from its use of transverse and longitudinal pulse position and momentum variances, the AG approach is applicable to both the single electron per pulse limit~\cite{lobastov_four-dimensional_2005}, where momentum variances determine the pulse evolution and the model is exact (obeying Gaussian optics), and the high charge density limit in which space-charge effects dominate~\cite{luiten_how_2004,siwick_ultrafast_2002,cao_femtosecond_2003}.
Unfortunately, the AG model as laid out by Michalik and Sipe lacks the ability to model the dynamics of electron pulses under influence of external forces; this prevents the AG model from being used to simulate the behavior of electron pulses in realistic microscope columns.
Additionally, to be useful for UEM, for which the initial electron pulse conditions are of notable importance for the performance of the instrument, a realistic set of intial conditions is essential for accurately modeling the pulse dynamics.

In Chapter \ref{chap:extension}, I present some optimizations to the AG model, and develop useful initial conditions representing single-photon photoemission from a typical metal photocathode.
I then present my extensions to the Michalik and Sipe AG model which permit the inclusion of generic linear external forces, such as those due to perfect electron-optical elements in a microscope column.
In order to adapt this extended AG model for simulating pulses in a realistic UEM column, I then present specific contributions that arise from magnetic lenses, RF cavities and DC accelerators.
The resulting computationally efficient propagation analysis can then be used to model, and hence design, UEM columns and UED systems in a straightforward manner.
These analyses presented are performed in the non-relativistic limit, which is a reasonable approximation for the typical 20-200keV electron energies employed in UED and electron microscopy.

In Chapter \ref{chap:model_results}, I apply the extended AG model to draw some initial insight into the dynamics of ultrafast electron pulses traversing a UEM column.
One of the conclusions that I will make is that oblate, or ``disk-like'', electron pulses will be generally preferred to prolate ``cigar-like.'' 
Indeed for UEM, with typical laser pulse durations in the femtosecond to picosecond timescales, the electron pulses are likely to be oblate for reasonably sized incident laser spot sizes.
Then in Chapter \ref{chap:considerations}, I apply some well known results from the field of electron microcopy, namely the Rose criterion~\cite{rose_television_1948}, an ultrafast analog to the Child-Langmuir law~\cite{child_discharge_1911,langmuir_effect_1923,valfells_effects_2002}, and the conservation of transverse emittance~\cite{jensen_emittance_2010} (each of which will be introduced) to arrive at two design goals for building a UEM which will produce high-resolution images.
First, that the electron pulses should be emitted from a large-area source, and second, that every effort must be made to reduce the initial rms transverse momentum (i.e., the emission cone angle) of the electron pulse as generated from the photocathode.

### 2013-04-12 main.tex
Some last-minute tweaks

\summary
Dynamic Transmission Electron Microscopy, and its picosecond/femtosecond subclass Ultrafast Electron Microscopy, is an emerging field in intrumentation science.
It attempts to combine the nanoscale spatial resolution of transmission electron microscopes with the temporal resolution of modern ultrafast lasers.
In this thesis I present my contributions to this young field. 
These include a novel model for simulating the dynamics of ultrafast electron pulses in electron microscope systems,
design critera for constructing such a system, and theoretical and experimental groundwork geared towards selecting a useful photocathode for electron generation.
I also present the prototype ultrafast electron microscope system being built at UIC.

>>>>

\summary
Dynamic Transmission Electron Microscopy, and its picosecond/femtosecond subclass Ultrafast Electron Microscopy, is an emerging field in intrumentation science.
It attempts to combine the nanoscale spatial resolution of transmission electron microscopes with the temporal resolution of modern ultrafast lasers.
In this thesis I present my contributions to this young field. 
These include a novel model for simulating the dynamics of ultrafast electron pulses in electron microscope systems,
design critera for constructing such a system, and theoretical and experimental groundwork geared towards selecting a useful photocathode for electron pulse generation.
I also present the prototype ultrafast electron microscope system being built at UIC.
